---
title: "What makes a Cinderella Team"
author: "Jacobi Lafferty"
date: "2026-02-2"
categories: [analysis, March Madness]
---

```{r}
library(tidyverse)
library(here)
```

#Introduction 
For this Post, I'm exploring March Madness Cinderella teams. I'm hoping to be able to answer the question of how to succesfully pick your Cinderella team when you are filling out your bracket. While it is true that it is mostly luck, and who gets hot at the right time, I'm hoping there are some underyling stats that will at least give us some guidance on what makes a good Cinderella team.

The dataset I'm working with is from kaggle called DEV _ MarchMadness.csv. You can find it here: https://www.kaggle.com/datasets/jonathanpilafas/2024-march-madness-statistical-analysis. It contains metrics for 8315 NCAA teams from 2002-2025. To answer my question, I needed to come up with some parameter's for what is considered a Cinderella team. I decided to have only 9 or above seeds in the tournament, and you had to make the final four. Originally I wanted to do elite 8 to allow for more teams to meet the criteria, but this dataset didn't have that. I would love to explore that further in the future. 

I wanted to examine both offense and defense to help answer my question. I chose 4 statistics from both sides of the ball, and hopefully a couple from each will prove to be significant. Offensively, I went with 3pt%, pace, TO%, and offensive rebounding. Defensively, I went with Foul rate, Opponent 3pt%, steal rate, and possession length. I thought out of the 165 statistics this data set offered, these 8 combined to cover most of the important aspects of the game. 

```{r}
# Read in The data 
DEV_March_Madness <- read_csv("/Users/jacobilafferty/Library/CloudStorage/OneDrive-St.LawrenceUniversity/DATA-334/ds334_blog/posts/Post1/archive/DEV _ March Madness.csv")
```

```{r}
# make the seed variable numeric
DEV_March_Madness <- DEV_March_Madness |> mutate(Seed = as.numeric(Seed))

# filter for only march madness teams and 9 or above seeds(to target Cinderella's)
DEV_March_Madness <- DEV_March_Madness |>
  filter(`Post-Season Tournament` == "March Madness",
         Seed >= 9
         )

# create a Cinderella variable
DEV_March_Madness <- DEV_March_Madness |>
  mutate(Cinderella = if_else(`Final Four?` == "Yes", "Yes", "No"))

```

# compare offense and defense
```{r}
## Plot comparing offensensive effeciency for Cinderella vs not Cinderella

ggplot(DEV_March_Madness, aes(x = Cinderella, y = AdjOE, fill = Cinderella)) +
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Offensive Efficiency", subtitle = "Points scored per 100 possesions", x = "Cinderella Team", y = "TO %") +
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```


```{r}
## Plot comparing defensive efficiency for Cinderella vs not Cinderella

ggplot(DEV_March_Madness, aes(x = Cinderella, y = AdjDE, fill = Cinderella)) +
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Defensive Efficiency", subtitle = "Points allowed per 100 possesions", x = "Cinderella Team", y = "TO %") +
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```


# Offense 


```{r}
# Boxplot comparing Pace for Cinderella vs not Cinderella
ggplot(DEV_March_Madness, aes(x = Cinderella, y = `Pre-Tournament.Tempo`, fill = Cinderella)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme_minimal() +
  labs(title = "Team Pace", subtitle = "# of possesions per 40 minutes", x = "Cinderella Team", y = "Pace") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```


```{r}
## BoxPlot comparing Rebounding for Cinderella vs not Cinderella

ggplot(DEV_March_Madness, aes(x = Cinderella, y = ORPct, fill = Cinderella)) +
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Team Offensive Rebounding", subtitle = "Percentage of own missed shots a team rebounded", x = "Cinderella Team", y = "Rebound %") +
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
## BoxPlot comparing Shooting for Cinderella vs not Cinderella

ggplot(DEV_March_Madness, aes(x = Cinderella, y = FG3Pct, fill = Cinderella)) +
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Team Three Point %", x = "Cinderella Team", y = "3pt %") +
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
## Plot comparing Turnovers for Cinderella vs not Cinderella

ggplot(DEV_March_Madness, aes(x = Cinderella, y = TOPct, fill = Cinderella)) +
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Team Turnover %", subtitle = "% of team's possesions that end in a turnover", x = "Cinderella Team", y = "TO %") +
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```



```{r}
#offensive scatterplot
library(ggrepel)

ggplot(data = DEV_March_Madness, aes(x = FG3Pct, y = TOPct)) +
  geom_point(data = subset(DEV_March_Madness, Cinderella == "No")) +
  geom_point(data = subset(DEV_March_Madness, Cinderella == "Yes"), color = "darkorange1", size = 3) +
  theme_minimal() +
  labs(title = "Cinderella Offensive Profile", 
       subtitle = "Comparing 3pt Accuracy and Ball Security",
       x = "3-Point Percentage", 
       y = "Turnover Percentage") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 
```

# defense 

```{r}
# Boxplot comparing Foul Rate for Cinderella vs not Cinderella
ggplot(DEV_March_Madness, aes(x = Cinderella, y = DefFT, fill = Cinderella)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme_minimal() +
  labs(title = "Defensive Foul Rate", 
       subtitle = "Free Throw Rate Allowed", 
       x = "Cinderella Team", y = "Foul Rate") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
# Boxplot comparing Steal Rate for Cinderella vs not Cinderella
ggplot(DEV_March_Madness, aes(x = Cinderella, y = StlRate, fill = Cinderella)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme_minimal() +
  labs(title = "Defensive Steal Rate", 
       subtitle = "Percentage of possessions ending in a steal", 
       x = "Cinderella Team", y = "Steal Rate") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
# Boxplot comparing Opponent 3pt % for Cinderella vs not Cinderella
ggplot(DEV_March_Madness, aes(x = Cinderella, y = OppFG3Pct, fill = Cinderella)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme_minimal() +
  labs(title = "Opponent 3Pt%", 
       x = "Cinderella Team", y = "Opp 3pt %") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
# Boxplot comparing Defensive Possession Length for Cinderella vs not Cinderella
ggplot(DEV_March_Madness, aes(x = Cinderella, y = `Avg Possession Length (Defense)`, fill = Cinderella)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("Yes" = "darkorange1", "No" = "darkgrey")) +
  theme_minimal() +
  labs(title = "Defensive Possesion Length", 
       subtitle = "Avg seconds per defensive possession", 
       x = "Cinderella Team", y = "Seconds") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
# defensive scatterplot 
  
ggplot(data = DEV_March_Madness, aes(x = OppFG3Pct, y = DefFT)) +
  geom_point(data = subset(DEV_March_Madness, Cinderella == "No")) +
  geom_point(data = subset(DEV_March_Madness, Cinderella == "Yes"), color = "darkorange1", size = 3) +
  theme_minimal() +
  labs(title = "Cinderella Defensive Profile", 
       subtitle = "Comparing Oppenent 3Pt% and Foul Rate",
       x = "Opponent 3-Point Percentage", 
       y = "Foul Rate") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

